Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-06-01 Manuscript/manuscript.tex
Clarity edits and supp lattice figures

\begin{abstract} 
Patterns of habitat quality are expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.  However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.  In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.  We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.  This suggests that for sufficiently long-lived environmental reservoirs, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.

Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infectious in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

We assume that patch population size is proportional to patch quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.  Building on the framework developed by \cite{Hanski2003} then,the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
\begin{align}
C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
\label{connectivity}
\end{align}
where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch population size; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

The model was parameterized  according to the following assumptions. (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).  (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).  To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity were varied over a range of possible values.

@@

For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$ and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for susceptible populations and $\nu A_i$ for infectious populations).

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).  To better isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_im$) and recorded the proportion of time spent occupied by infectious hosts for each patch (as a measure of disease risk) and the final total effective population size of the host.

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was determined largely by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) allowing pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).

@@

For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
In particular, for a given disease, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).  
For a given pathogen in this range (e.g. with $\nu = 0.2$ and $\delta = 0.3$), this relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the relative increase in the persistence and spread of environmentally long-lived pathogens in high quality habitat distributions.

To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
However, disease risk, measured by the probability of infection, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs. d-f).
Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially (Fig~\ref{pext}).

Blah blah blah....something about lattice looking similar (supp info).

>>>>

\begin{abstract} 
The distribution of habitat quality is expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.  However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.  In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.  We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.  This suggests that for sufficiently long-lived environmental reservoirs, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes in turn may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

We assume that patch population size is proportional to patch quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.  Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
\begin{align}
C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
\label{connectivity}
\end{align}
where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

The model was parameterized  according to the following assumptions (Table~\ref{params}). (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).  To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity were varied over a range of possible values.

@@

For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for susceptible populations and $\nu A_i$ for infectious populations).

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk) and the final total effective host population size.

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was determined largely by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

@@

For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
In particular, for a given disease, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).  
For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions.
Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).

To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
However, disease risk, measured by the proportion of time a patch is occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  
Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e. $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs. d-f).
Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e. $\xi_{em} = 0$, Fig~\ref{pext}).
